Simulating the Lipkin-Meshkov-Glikov model on a classical computer

### 2023-06-09 doc/paper.tex
Fixing som typos
Signed-off-by: Gregz9 <gregorkajda@gmail.com>

\title{Simulating the Lipkin-Meshov-Glikov model on a classical computer}

@@

We present a comprehensive simulation of the simplified Hamiltonian of the Lipkin-Meshov-Glikov (LMG) model on a classical computer, employing the Variational Quantum Eigensolver (VQE) algorithm. Initially, we explore the outcomes arising from the application of diverse single qubit quantum gates to one qubit systems prepared in basis states. Subsequently, an entangler circuit is constructed, and one of the four bell pairs is prepared, to which we apply the Hadamard and CNOT gates. Utilizing our newfound grasp of the foundational principles governing quantum systems, we construct quantum circuits for generating ansatz states and methodically develop the VQE from ground up.

To assess the VQE's efficacy, we apply it to quantum systems representing single and two-particle systems, comparing its performance against an analytical solver. Transitioning to the Lipkin model, we leverage the properties of annihilation and creation operators to reformulate the Lipkin model's Hamiltonian first in terms of the quasi-spin operator and then Pauli operators. Employing this encoding scheme, we apply the VQE to systems with spin J=1 and 2, corresponding to Lipkin systems with two and four fermions, respectively. We compare the results against exact solutions obtained through standard diagonalization methods. This analysis provides valuable insights into the performance and accuracy of the VQE approach in tackling the Lipkin model's Hamiltonian.

@@

An emerging research area in quantum computing focuses on the simulation of nuclear physics, leveraging the unique properties of quantum systems. Nuclear systems possess distinct characteristics that make them well-suited for exploration using quantum computers, providing insights into particle interactions. However, it is worth noting that quantum computers still face challenges, such as larger error rates compared to classical computers and vulnerability to noise, which can make working with them potentially difficult. Although the field of quantum computing is expanding rapidly, widespread accessibility remains limited, with companies like IBM offering only highly restricted access to their quantum technology. In light of these factors, we will employ a classical computer to simulate and analyze the simplified Lipkin-Meshkov-Glick model, showcasing the potential that quantum computing holds while acknowledging the current limitations in accessibility.

@@

 \subsection{Quantum gates}
The primary distinction between classical computers and their quantum counterparts lies in the nature of the data they process. In classical computing, the fundamental data unit is the binary digit, commonly known as a bit, which represents different voltage levels associated with low and high states. On the other hand, quantum computers utilize particles, such as atoms, as the information unitm leading to a fundamentally different approach to data handling. The unique properties of atoms and the underlying principles of quantum mechanics prevent us from treating them in the same manner as classical bits. For instance, in classical computing, logical gates like \textit{and}, \textit{or}, \textit{xor}, and \textit{nand} are commonly used for data processing, but these gates produce irreversible outcomes. However, on quantum computers, irreversible outcomes are generally not allowed, except in the case of measurement.

On a quantum computer, the fundamental data type used for storing information is the \textit{qubit}, which represents the digital manifestation of the wave function for a quantum particle. The principles of quantum mechanics dictate that the evolution of a quantum system can be mathematically described using unitary operators. This implies that the operations applied to quantum systems are inherently reversible. Consequently, at any given point in time, a quantum system can be fully characterized by a set of linear unitary operators. As a result, data processing on a quantum computer necessitates the use of reversible operators, enabling the "uncomputation" of the outcomes produced by quantum gates. The most basic of these operators are typically represented using 2x2 matrices. While these 2x2 operators quantum gates operate on individual qubits, by utilizing the Kronecker product (tensor product), we can in theory construct gates for quantum systems composed of arbitrary number of qubits. In practice we will often find ourselves restricted to at most 8 or 10 qubits before the computational overhead becomes too high.

@@

	\item[\textbf{III.}] \textbf{Pauli-Y gate} \\
		The action of the Pauli Y gate can be described as the mapping of $|0>$ to $i|1>$ and $|1>$ to $i¦0>$. The action of the gate can also be viewed as a rotation of the state of the qubit around the y-axis of the Bloch sphere by 180 degrees. Presented below is the matrix representation of the quantum gate the result of applying it to a single qubit:

@@

	\item[\textbf{IV.}] \textbf{Pauli-Z gate} \\
		The Pauli Z gate flips the phase of $|1>$ to $-|1>$ while leaving $|0>$, meaning it flips the relative phase of the qubit. Presented below is the matrix representation of the quantum gate the result of applying it to a single qubit:

@@

	\item[\textbf{VII.}] \textbf{CNOT-gate} \\
		The Controlled-NOT gate, often abbreviated as CNOT, is a more complex gate that combines the Identity and Pauli-X gates using a Kronecker product. This gate is designed to operate on quantum systems composed of two qubits rather than a single qubit. The name of the gate provides a clue about its functionality. By using the first qubit as a control signal, the CNOT gate flips the amplitudes of the second qubit if the control qubit is in the basis state |1>, which corresponds to an active high signal in classical computing. However, if the control qubit is in the state |0>, no changes are applied to the second qubit. The specific gate we are discussing here is the $CNOT_{0,1}$-gate, where the first qubit serves as the control signal.
		It's important to note that we can also use the second qubit as the control signal by simply changing the order of the gates in the tensor product used to create the CNOT gate from $I \otimes X$ to $X \otimes I$. In matrix form, this operator is represented by a 4x4 matrix that can be written as:

@@

By utilizing quasi-spin operators, we can account for these internal degrees of freedom and incorporate them into our calculations. This allows us to rewrite the Hamiltonian of the Lipkin model in a more convenient form. Consequently, we can express the Hamiltonian matrix in terms of the Pauli operators mentioned in the previous subsection. This transformation facilitates the analysis and study of the Lipkin model, providing insights into its energy levels and dynamics. Before proceeding however, we are going to simplify the Hamiltonian slightly more by setting the spin-exchange term equal to zero, resulting in a dimensionless Hamiltonian
\begin{equation}
\hat{H} = \epsilon\hat{J}_z - \frac{1}{2}V(\hat{J}_+\hat{J}_+ + \hat{J}_-\hat{J}_-).
\end{equation}

>>>>

\title{Simulating the Lipkin-Meshkov-Glikov model on a classical computer}

@@

We present a comprehensive simulation of the simplified Hamiltonian of the Lipkin-Meshov-Glikov (LMG) model on a classical computer, employing the Variational Quantum Eigensolver (VQE) algorithm. Initially, we explore the outcomes arising from the application of diverse single qubit quantum gates to one qubit systems prepared in basis states. Following that, an entangler circuit is constructed, and one of the four bell pairs is prepared, to which we apply the Hadamard and CNOT gates. Utilizing our newfound grasp of the foundational principles governing quantum systems, we construct quantum circuits for generating ansatz states and methodically develop the VQE from ground up.

To assess the VQE's efficiency, we apply it to quantum systems representing single and two-particle systems, comparing its performance against an analytical solver. Transitioning to the Lipkin model, we leverage the properties of annihilation and creation operators to reformulate the Lipkin model's Hamiltonian first in terms of the quasi-spin operator and then Pauli operators. Employing this encoding scheme, we apply the VQE to the Lipkin model with spin J=1 and 2, corresponding to situation when the model comprises two and four fermions, respectively. We compare the results against exact solutions obtained through standard diagonalization method. This analysis provides valuable insights into the performance and accuracy of the VQE approach in tackling the Lipkin model's Hamiltonian.

@@

An emerging research area in quantum computing focuses on the simulation of nuclear physics, leveraging the unique properties of quantum systems. Nuclear systems possess distinct characteristics that make them well-suited for exploration using quantum computers, providing insights into particle interactions. However, it is worth noting that quantum computers still face challenges, such as larger error rates compared to classical computers and vulnerability to noise, which can make working with them potentially difficult. Although the field of quantum computing is expanding rapidly, widespread accessibility remains limited, with companies like IBM offering only restricted access to their quantum technology. In light of these factors, we will employ a classical computer to simulate and analyze the simplified Lipkin-Meshkov-Glick model, showcasing the potential that quantum computing holds while acknowledging the current limitations in accessibility.

@@

 \subsection{Quantum gates}
The primary distinction between classical computers and their quantum counterparts lies in the nature of the data they process. On a classical computer, the fundamental data unit is the binary digit, commonly known as a bit, which represents different voltage levels associated with low and high states, or 0 and 1 as is also common.Quantum computers on the other hand, utilize particles as the information unit, hence leading to a fundamentally different approach to data handling. The unique properties of atoms and the underlying principles of quantum mechanics prevent us from treating them in the same manner as classical bits. For instance, in classical computing, logical gates like \textit{and}, \textit{or}, \textit{xor}, and \textit{nand} are commonly used for data processing, but these gates produce irreversible outcomes. However, on quantum computers, irreversible outcomes are generally not allowed, except in the case of measurement.

On a quantum computer, the fundamental data type used for storing information is the \textit{qubit}, which represents the digital manifestation of the wave function for a quantum particle. The principles of quantum mechanics dictate that the evolution of a quantum system can be mathematically described using unitary operators. This implies that the operations applied to quantum systems are inherently reversible. As a result, data processing on a quantum computer necessitates the use of reversible operators, enabling the "uncomputation" of the outcomes produced by quantum gates. The most basic of these operators are typically represented using 2x2 matrices. While these 2x2 operators quantum gates operate on individual qubits, by utilizing the Kronecker product (tensor product), we can in theory construct gates for quantum systems composed of arbitrary number of qubits. In practice we will often find ourselves restricted to at most 8 or 10 qubits before the computational overhead becomes too high.

@@

	\item[\textbf{III.}] \textbf{Pauli-Y gate} \\
		The action of the Pauli Y gate can be described as the mapping of $|0>$ to $i|1>$ and $|1>$ to $i¦0>$. The action of the gate can also be viewed as a rotation of the state of the qubit around the y-axis of the Bloch sphere (\textit{see Fig.1}) by 180 degrees. Presented below is the matrix representation of the quantum gate the result of applying it to a single qubit:

@@

	\item[\textbf{IV.}] \textbf{Pauli-Z gate} \\
		The Pauli Z gate flips the phase of $|1>$ to $-|1>$ while leaving $|0>$ unchanged, meaning it flips the relative phase of the qubit. Presented below is the matrix representation of the quantum gate the result of applying it to a single qubit:

@@

	\item[\textbf{VII.}] \textbf{CNOT-gate} \\
		The Controlled-NOT gate, often abbreviated as CNOT, is a more complex gate that combines the Identity and Pauli-X gates using a Kronecker product. This gate is designed to operate on quantum systems composed of two qubits rather than a single qubit. The name of the gate provides a clue about its functionality. By using the first qubit as a control signal, the CNOT gate flips the amplitudes of the second qubit if the control qubit is in the basis state $|1>$, which corresponds to an active high signal in classical computing. However, if the control qubit is in the state $|0>$, no changes are applied to the second qubit. The specific gate we are discussing here is the $CNOT_{0,1}$-gate, where the first qubit serves as the control signal.
		It's important to note that we can also use the second qubit as the control signal by simply changing the order of the gates in the tensor product used to create the CNOT gate from $I \otimes X$ to $X \otimes I$. In matrix form, this operator is represented by a 4x4 matrix that can be written as:

@@

By utilizing quasi-spin operators, we can account for these internal degrees of freedom and incorporate them into our calculations. This allows us to rewrite the Hamiltonian of the Lipkin model in a more convenient form. Consequently, we can express the Hamiltonian matrix in terms of the Pauli operators mentioned in the previous subsection, just as we did with the one- and two-qubit Hamiltonian. This transformation facilitates the analysis and study of the Lipkin model, providing insights into its energy levels and dynamics. Before proceeding however, we are going to simplify the Hamiltonian slightly more by setting the spin-exchange term equal to zero, resulting in a dimensionless Hamiltonian
\begin{equation}
\hat{H} = \epsilon\hat{J}_z - \frac{1}{2}V(\hat{J}_+\hat{J}_+ + \hat{J}_-\hat{J}_-).
\end{equation}
